Performance-portable Parallel Programming in Computer Science Education with Kokkos

### 2020-01-24 src/1.Introduction.tex
Adding other example and back-end example.

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed through a multi-institutional collaboration between US national laboratories as part of the US Department of Energy Exascale Project\cite{ECP}. It consists of a programming model specification and a library. The specification defines execution and memory model and an application programming interface (library API). The runtime library exposes programming primitives, namely, patterns and types and maps them to the underlying abstract machine model. The machine model creates a generic view of compute hardware and memories. The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts. Kokkos has been successfully used in many large-scale projects at SNL, Utah University, ORNL, LANL and other research centers~\cite{KOKKOSUSECASE2}. We list its key properties as follows.

>>>>

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed collaboratively between national laboratories as part of the US Department of Energy Exascale Project\cite{ECP}. It consists of a programming model specification and a library. The specification defines execution and memory model and an application programming interface (library API). The runtime library exposes programming primitives, namely, patterns and types and maps them to the underlying abstract machine model. The machine model creates a generic view of compute hardware and memories. The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts. Kokkos has been successfully used in many large-scale projects at SNL, University of Utah, Oak-Ridge National Laboratories, Los Alamos National Laboratories and others~\cite{KOKKOSUSECASE2}. We list its key properties as follows.

### 2020-01-24 src/5.Example.tex
Renamed files for consistency.
Small changes.
Adding other example and back-end example.

\section{Back-end Support}\label{chap:kokkosBackend}

This section provides a practical insight into programming with Kokkos and the inner workings of the programming model library. Figure~\ref{fig:KokkosExample} shows am example application that algorithmically initializes an array to integers and includes several points of interests. Taking a closer look, the reader can observe the initialization and finalization of the Kokkos runtime library through the~\emph{Kokkos::initialize} and~\emph{Kokkos::finalize} method calls. Further the example shows the use of the~\emph{view} abstraction that represents a one-dimensional integer array of size~\emph{N}. The object carries the name~\emph{A} internally to facilitate tracing and debugging. In this example, a parallel\_for pattern object receives a range policy that is templated to the OpenMP execution space and takes~\emph{N} and a lambda expression as parameters. Vies behave as pointers and there a copy-semantic in the lambda expression is used in this case. Within the loop-body, the view object is accessed through the~\emph{()-operator}. It is up to the OpenMP execution space to provide performance implementations of the view, an efficient internal representation of the array and for the access operator for the underlying hardware architecture.

@@

\begin{figure}
\begin{small}
\begin{Verbatim}[frame=leftline]
int main(int argc, char* argv[]) {
  Kokkos::initialize(argc,argv);
  {
    int N = atoi(argv[1]);
    Kokkos::View<int*> a("A", N);  
    Kokkos::parallel_for(
    Kokkos::RangePolicy<Kokkos::OpenMP>(N,
    [=](const int & i) {
      a(i)=i*i;
    }); 
  }
  Kokkos::finalize();
}
\end{Verbatim}
\end{small}
\caption{This example application written in Kokkos shows the use of a~\emph{view}-type and a parallel for loop. Both Kokkos types are templated to the OpenMP execution space.}
\label{fig:KokkosExample}
\end{figure}

>>>>

\section{Example}\label{chap:kokkosExample}

This section provides a practical insight into programming with Kokkos and the inner workings of the programming model library. Figure~\ref{fig:KokkosExample} shows am example application that algorithmically initializes an array to integers. Taking a closer look, the reader can observe the initialization and finalization of the Kokkos runtime library through the~\emph{Kokkos::initialize} and~\emph{Kokkos::finalize} method calls. Further the example shows the use of the~\emph{view} abstraction that represents a one-dimensional integer array of size~\emph{N}. The object carries the name~\emph{A} internally to facilitate tracing and debugging. In this example, a parallel\_for pattern object receives a range policy that is templated to the OpenMP execution space and takes~\emph{N} and a lambda expression as parameters. Vies behave as pointers and there a copy-semantic in the lambda expression is used in this case. Within the loop-body, the view object is accessed through the~\emph{()-operator}. It is up to the OpenMP execution space to provide performance implementations of the view, an efficient internal representation of the array and for the access operator for the underlying hardware architecture.

@@

\begin{figure}
\begin{small}
\begin{Verbatim}[frame=leftline]
int main(int argc, char* argv[]) {
  Kokkos::initialize(argc,argv);
  {
    ...
    Kokkos::View<double*> a("A", dN);  
    Kokkos::View<double*> r("A", N);
    Kokkos::View<double**> M("A", dM, dN);

    Kokkos::parallel_for("outer", TeamPolicy<>(M,AUTO),
    [=](const member_type &teamID)
    {
      const int e = teamID.league_rank();
      Kokkos::parallel_reduce( TeamThreadRange(teamID, N),
        [=](const int & i, double & r_tmp) {
          r_tmp += M(e, i) * a(i);
        }, 
        r(e));
    }); 
  }
  Kokkos::finalize();
}
\end{Verbatim}
\end{small}
\caption{This example application written in Kokkos shows the use of a~\emph{view}-type and a parallel for loop. Both Kokkos types are templated to the OpenMP execution space.}
\label{fig:KokkosExample}
\end{figure}

%\begin{figure}
%\begin{small}
%\begin{Verbatim}[frame=leftline]
%int main(int argc, char* argv[]) {
%  Kokkos::initialize(argc,argv);
%  {
%    int N = atoi(argv[1]);
%    Kokkos::View<int*> a("A", N);  
%    Kokkos::parallel_for(
%    Kokkos::RangePolicy<Kokkos::OpenMP>(N,
%    [=](const int & i) {
%      a(i)=i*i;
%    }); 
%  }
%  Kokkos::finalize();
%}
%\end{Verbatim}
%\end{small}
%\caption{This example application written in Kokkos shows the use of a~\emph{view}-type and a parallel for loop. Both Kokkos types are templated to the OpenMP execution space.}
%\label{fig:KokkosExample}
%\end{figure}

### 2020-01-24 src/9.Conclusion.tex
Adding other example and back-end example.

\section{Conclusion}\label{chap:conclusion}
%...
%Consequently, we emphasize the importance and welcome new development and educational use of parallel programming models that align to this trend. We believe that abstracting from vendor or architecture-specific programming models not only simplifies parallel programming in science and industry but also would support educators in teaching parallel programming.
%We would also like to express gratitude to educators for taking on this important role. Further materials, including a detailed description of the programming- and machine model can be found by following the reference on the Kokkos programming mode. Finnaly, we would like to mention that interested developers and student can reach the development team on Slack by following the invititation link as mentioned in the reference section.

>>>>

\section{Conclusion}\label{chap:conclusion}
%...
In this work we gave an introduction to the Kokkos programming model. We presented a conceptual framework consi,  presented the Kokkos parallel prgroamming model. 
Programming support for parallel progrmaming in mainstream langiages is gaining momentung
. Programming abstraction libraries such as Kokkos, Raja and others showcase this trend. Targeting a generic parallel machine representation allows to design and efficiently use

We emphasize the importance and welcome new development and educational use of parallel programming models that align to this trend. We believe that abstracting from vendor or architecture-specific programming models not only simplifies parallel programming in science and industry but also would support educators in teaching parallel programming.
We would also like to express gratitude to educators for taking on this important role. Further materials, including a detailed description of the programming- and machine model can be found by following the reference on the Kokkos programming mode. Finally, we would like to mention that interested developers and student can reach the development team on Slack by following the invitation link as mentioned in the reference section.
